Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-10-28 paper/paper.tex
flip linear arrows

% TODO this should be beter formatted but can't use listings with \otimes ^{\ast}
\begin{figure}
\begin{align*}
\data &~A^{*}: L_{i}~\where\\
      & \nil : I \lto A^{*} \\
      & \cons : A \otimes A^{*} \lto A^{*}
\end{align*}
\caption{Kleene Star as an inductive type}
\label{fig:kleenestarinductive}
\end{figure}

@@

@@

\begin{figure}
\begin{align*}
\data &~\Trace_{N}~1~(acc : \Bool) : L_{i}~\where\\
      & \mathsf{stop}_{1} : I \lto \Trace_{N}~1~\true \\
      & \mathsf{1to2} : \forall (acc : \Bool) \to b \otimes \Trace_{N}~2~acc \lto \Trace_{N}~1~acc \\
      & \mathsf{1to3} : \forall (acc : \Bool) \to \Trace_{N}~3~acc \lto \Trace_{N}~1~acc \\
\\
\data &~\Trace_{N}~2~(acc : \Bool) : L_{i}~\where\\
      & \mathsf{stop}_{2} : I \lto \Trace_{N}~2~\false \\
      & \mathsf{2to2} : \forall (acc : \Bool) \to a \otimes \Trace_{N}~2~acc \lto \Trace_{N}~2~acc \\
      & \mathsf{2to3} : \forall (acc : \Bool) \to b \otimes \Trace_{N}~3~acc \lto \Trace_{N}~2~acc \\
\\
\data &~\Trace_{N}~3~(acc : \Bool) : L_{i}~\where\\
      & \mathsf{stop}_{3} : I \lto \Trace_{N}~3~\true \\
      & \mathsf{3to1} : \forall (acc : \Bool) \to a \otimes \Trace_{N}~1~acc \lto \Trace_{N}~3~acc \\
\end{align*}
\caption{$N$ as an indexed inductive type}
\label{fig:nfainductive}
\end{figure}

@@

\begin{figure}
  \[
  \begin{array}{rrcl}
    \textrm{non-linear contexts} & \Gamma & ::= & \cdot \pipe \Gamma,x:X \\
    \textrm{non-linear types} & X , Y , Z & ::= & \mathsf{Nat} \pipe
                                                  \mathsf{Int} \pipe
                                                  \mathsf{Empty} \pipe
                                                  \mathsf{Unit} \pipe
                                                  \mathsf{SPF}\,X\pipe U \pipe L
                                                  \pipe \ltonl A \pipe \sum_{x:X} Y
                                                  \pipe \prod_{x:X} Y\pipe M =_A
                                                  N \\
    \textrm{linear contexts} & \Delta & ::= & \cdot \pipe \Delta , a:A\\
    \textrm{linear types} & A , B , C & ::= &
    c \pipe I \pipe A \otimes B \pipe \bigoplus\limits_{x:X} A \pipe \mu A_f M
    A \lto B \pipe B \tol A \pipe \bigwith\limits_{x:X} A \pipe \equalizer {a}{f}{g}\\
  \end{array}
  \]
  \caption{Overview of Types}
\end{figure}

@@

    \inferrule{~}{\Gamma ; a : A \vdash a : A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
    %
    \\
    %
    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : I \\ \Gamma ; \Delta_1',\Delta_2' \vdash e' : C}{\Gamma ; \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
    %
    \\
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
    \\
    %
    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
    \and
    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta', \Delta \vdash \applto {e'} {e} : B}
    \\
    %
    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta, \Delta' \vdash \apptol e {e'} : B}
    %
    \\
    %
    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}
              {\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash e' : X}{\Gamma ; \Delta \vdash e\,.\pi\,M : \subst A {e'} x}
    %
    \\
    %
    \inferrule{\Gamma \vdash M : X \quad \Gamma ; \Delta \vdash e : \subst A e x}{\Gamma ; \Delta \vdash \sigma\,M\,e : \bigoplus\limits_{x:X} A}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta \vdash e : \bigoplus\limits_{x:X} A \quad \Gamma, x : X ; \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma; \Delta'_1, \Delta, \Delta'_2 \vdash \letin {\sigma\,x\,a} e {e'}: C}
    %
    \and
    %
    \inferrule
    {~}
    {\Gamma , x : \ltonl A ; \cdot \vdash x : A}
    %
    \and
    %
    \inferrule
    {\Gamma \vdash f : \ltonl {(A \lto B)} \\
      \Gamma \vdash g : \ltonl { (A \lto B) } \\
      \Gamma ; \Delta \vdash a : A \\
      \Gamma ; \Delta \vdash p : \apptol {f}{a} = \apptol {g}{a}}
    {\Gamma ; \Delta \vdash \equalizerpair{a}{p} : \equalizer {a}{f}{g}}
    %
    \and
    %
    \inferrule
    {\Gamma \vdash f : \ltonl { (A \lto B) } \\
      \Gamma \vdash g : \ltonl { (A \lto B) } \\
      \Gamma ; \Delta \vdash e : \equalizer{a}{f}{g}}
    {\Gamma ; \Delta \vdash \equalizerpi {e} : A}
    %% \\
    %% %
    %% \inferrule{~}{\Gamma ; \Delta \vdash () : \top}

@@

    % \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
%
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e x {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e x {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : A \tol B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {e'} \equiv \subst x a {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} a e : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {e \otimes e'} {a \otimes a'} e'' \equiv \subst {e''} {a, a'} {e, e'} : C}
%
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} c e : C}
%
    \and
%
    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(e, e')} {e''} \equiv \subst {e''} {x, a} {e, e'} : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
    \and
    %
    \inferrule
    {\Gamma \vdash f : \ltonl { (A \lto B) } \\
      \Gamma \vdash g : \ltonl { (A \lto B) } \\
      \Gamma ; \Delta \vdash a : A \\
      \Gamma ; \Delta \vdash p : \apptol {f}{a} \equiv \apptol {g}{a}}
    {\Gamma ; \Delta \vdash \equalizerpi {\equalizerpair {a}{p}} \equiv a : A}
    %
    \and
    %
    \inferrule
    {\Gamma \vdash f : \ltonl { (A \lto B) } \\
      \Gamma \vdash g : \ltonl { (A \lto B) } \\
      \Gamma ; \Delta \vdash p : \apptol {f}{a} = \apptol {g}{a} \\
      \Gamma ; \Delta \vdash e : \equalizer{a}{f}{g}}
    {\Gamma ; \Delta \vdash \equalizerpair {\equalizerpi {e}} {p} \equiv e : A}
\end{mathpar}
  \caption{Judgmental equality}
  \label{fig:jdgeq}
\end{figure}
\steven{Do we need inverse for $\ltonl A$ in judgemental equality figure?}
\steven{Do we need to say anything about the equalizer proofs being unique in
  judgemental equality figure?}

>>>>

% TODO this should be beter formatted but can't use listings with \otimes ^{\ast}
\begin{figure}
\begin{align*}
\data &~A^{*}: L_{i}~\where\\
      & \nil : \ltonl {A^{*}} \\
      & \cons : \ltonl {(A \lto A^{*} \lto A^{*})}
\end{align*}
\caption{Kleene Star as an inductive type}
\label{fig:kleenestarinductive}
\end{figure}

@@

\steven{TODO need to update the proof tree example derivations (\cref{fig:kleenestarderivation,fig:kleeneabstractproof}) to use the presentation of cons that is now in \cref{fig:kleenestarinductive}}

@@

\begin{figure}
\begin{align*}
\data &~\Trace_{N}~1~(acc : \Bool) : L_{i}~\where\\
      & \mathsf{stop}_{1} : \ltonl {\Trace_{N}~1~\true}  \\
      & \mathsf{1to2} : \forall (acc : \Bool) \to \ltonl {(b \lto \Trace_{N}~2~acc \lto \Trace_{N}~1~acc)} \\
      & \mathsf{1to3} : \forall (acc : \Bool) \to \ltonl {(\Trace_{N}~3~acc \lto \Trace_{N}~1~acc)} \\
\\
\data &~\Trace_{N}~2~(acc : \Bool) : L_{i}~\where\\
      & \mathsf{stop}_{2} : \ltonol {\Trace_{N}~2~\false} \\
      & \mathsf{2to2} : \forall (acc : \Bool) \to \ltonl {(a \lto \Trace_{N}~2~acc \lto \Trace_{N}~2~acc)} \\
      & \mathsf{2to3} : \forall (acc : \Bool) \to \ltonl {(b \lto \Trace_{N}~3~acc \lto \Trace_{N}~2~acc)}  \\
\\
\data &~\Trace_{N}~3~(acc : \Bool) : L_{i}~\where\\
      & \mathsf{stop}_{3} : \ltonl {\Trace_{N}~3~\true} \\
      & \mathsf{3to1} : \forall (acc : \Bool) \to \ltonl {(a \lto \Trace_{N}~1~acc \lto \Trace_{N}~3~acc)} \\
\end{align*}
\caption{$N$ as an indexed inductive type}
\label{fig:nfainductive}
\end{figure}

@@

\begin{figure}
  \[
  \begin{array}{rrcl}
    \textrm{non-linear contexts} & \Gamma & ::= & \cdot \pipe \Gamma,x:X \\
    \textrm{non-linear types} & X , Y , Z & ::= & \mathsf{Nat} \pipe
                                                  \mathsf{Int} \pipe
                                                  \mathsf{Empty} \pipe
                                                  \mathsf{Unit} \pipe
                                                  \mathsf{SPF}\,X\pipe U \pipe L
                                                  \pipe \ltonl A \pipe \sum_{x:X} Y
                                                  \pipe \prod_{x:X} Y\pipe M =_A
                                                  N \\
    \textrm{linear contexts} & \Delta & ::= & \cdot \pipe \Delta , a:A\\
    \textrm{linear types} & A , B , C & ::= &
    c \pipe I \pipe A \otimes B \pipe \bigoplus\limits_{x:X} A \pipe \mu A_f M \pipe
    A \lto B \pipe B \tol A \pipe \bigwith\limits_{x:X} A \pipe \equalizer {a}{f}{g}\\
  \end{array}
  \]
  \caption{Overview of Types}
\end{figure}

@@

    \inferrule{~}{\Gamma ; a : A \vdash a : A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
    %
    \\
    %
    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : I \\ \Gamma ; \Delta_1',\Delta_2' \vdash e' : C}{\Gamma ; \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
    %
    \\
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
    \\
    %
    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
    \and
    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta, \Delta' \vdash \applto {e'} {e} : B}
    \\
    %
    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta', \Delta \vdash \apptol e {e'} : B}
    %
    \\
    %
    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}
              {\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash e' : X}{\Gamma ; \Delta \vdash e\,.\pi\,M : \subst A {e'} x}
    %
    \\
    %
    \inferrule{\Gamma \vdash M : X \quad \Gamma ; \Delta \vdash e : \subst A e x}{\Gamma ; \Delta \vdash \sigma\,M\,e : \bigoplus\limits_{x:X} A}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta \vdash e : \bigoplus\limits_{x:X} A \quad \Gamma, x : X ; \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma; \Delta'_1, \Delta, \Delta'_2 \vdash \letin {\sigma\,x\,a} e {e'}: C}
    %
    \and
    %
    \inferrule
    {~}
    {\Gamma , x : \ltonl A ; \cdot \vdash x : A}
    %
    \and
    %
    \inferrule
    {\Gamma \vdash f : \ltonl {(A \lto B)} \\
      \Gamma \vdash g : \ltonl { (A \lto B) } \\
      \Gamma ; \Delta \vdash a : A \\
      \Gamma ; \Delta \vdash p : \apptol {f}{a} = \apptol {g}{a}}
    {\Gamma ; \Delta \vdash \equalizerpair{a}{p} : \equalizer {a}{f}{g}}
    %
    \and
    %
    \inferrule
    {\Gamma \vdash f : \ltonl { (A \lto B) } \\
      \Gamma \vdash g : \ltonl { (A \lto B) } \\
      \Gamma ; \Delta \vdash e : \equalizer{a}{f}{g}}
    {\Gamma ; \Delta \vdash \equalizerpi {e} : A}
    %% \\
    %% %
    %% \inferrule{~}{\Gamma ; \Delta \vdash () : \top}

@@

    % \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
%
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e x {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e x {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : B \tol A}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {e'} \equiv \subst x a {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} a e : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {e \otimes e'} {a \otimes a'} e'' \equiv \subst {e''} {a, a'} {e, e'} : C}
%
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} c e : C}
%
    \and
%
    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(e, e')} {e''} \equiv \subst {e''} {x, a} {e, e'} : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
    \and
    %
    \inferrule
    {\Gamma \vdash f : \ltonl { (A \lto B) } \\
      \Gamma \vdash g : \ltonl { (A \lto B) } \\
      \Gamma ; \Delta \vdash a : A \\
      \Gamma ; \Delta \vdash p : \apptol {f}{a} \equiv \apptol {g}{a}}
    {\Gamma ; \Delta \vdash \equalizerpi {\equalizerpair {a}{p}} \equiv a : A}
    %
    \and
    %
    \inferrule
    {\Gamma \vdash f : \ltonl { (A \lto B) } \\
      \Gamma \vdash g : \ltonl { (A \lto B) } \\
      \Gamma ; \Delta \vdash p : \apptol {f}{a} = \apptol {g}{a} \\
      \Gamma ; \Delta \vdash e : \equalizer{a}{f}{g}}
    {\Gamma ; \Delta \vdash \equalizerpair {\equalizerpi {e}} {p} \equiv e : A}
\end{mathpar}
  \caption{Judgmental equality}
  \label{fig:jdgeq}
\end{figure}
\steven{Do we need inverse for $\ltonl A$ in judgemental equality figure?}
\steven{Do we need to say anything about the equalizer proofs being unique in
  judgemental equality figure?}
